Human Factors Influencing Visual Statistical Inference

### 2013-06-06 sociology_paper.tex
one more tweak to abstract
performance plots by demographics added

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get better and faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement of actual data plot in the lineups does not affect the inference.

@@

>>>>

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get better and faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement of actual data plot in the lineups does not affect the inference.

@@

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=6.5in]{age_gender_within_country_time.pdf} 
   \caption{Countrywise average time taken for different age and academic levels of the MTurk workers participating the experiments shows that the demographic factors may not have effect on time taken. }
   \label{fig:gender_country}
\end{figure}

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=6.5in]{age_gender_within_country_correct.pdf} 
   \caption{Countrywise percentage of correct responses for different age and academic levels of the MTurk workers participating the experiments shows that the demographic factors may not have effect on the percentage of correct responses. }
   \label{fig:gender_country}
\end{figure}
